---
title: "HW2 Q1"
author: "Vy Dang"
date: "2024-09-19"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

### Problem 1 (Through lecture 6)
We have data from a simple random sample of 300 work-force participants born before 1997 in a rural area. Using this data, we will try to understand the relationship between education and income within this
area. Among the variables at our disposal are the years of education for an individual, stored as Education,
and what their income is (in thousands of dollars), stored as Income. The data is stored in the data set
edu.csv. In the questions that follow, you may assume that the requisite assumptions needed constructing
confidence intervals and performing hypothesis tests using R’s multiple regression output hold.

a. (2 pts) Provide a scatterplot with Income on the y-axis and Education on the x. Discuss whether the
relationship seems linear, and whether the association appears to be positive or negative.
```{r}
edu <- read.csv("edu.csv")
```

```{r}
ggplot(edu, aes(y = Income, x = Education)) + geom_point()
```

Given that the scatterplot shows vertical lines for each education level without forming any discernible linear pattern, the relationship between Education and Income might not be linear. There seems to be positive association here in the graph. 

b. (2 pts) Run a simple regression of Income on Education, and show the resulting prediction equation.
We will refer to this analysis as Analysis I moving forwards.

```{r}
Analysis_I <- lm(Income ~ Education, data = edu)
summary(Analysis_I)
```
The prediction equation: Income = 25.2100 - 0.5172 * Education

c. (4 pts) Is there evidence that the slope coefficient on Education is less than -0.1? Conduct your test
at α = 0.05.

Ho: $\beta_1 \geq -0.1$
Ha: $\beta_1 < -0.1$

```{r}
coefficients <- coef(summary(Analysis_I))
beta1_hat <- coefficients["Education", "Estimate"]
se_beta1_hat <- coefficients["Education", "Std. Error"]
tstat <- (beta1_hat - (-0.1)) / se_beta1_hat
pvalue <- pt(tstat, df = Analysis_I$df.residual, lower.tail = TRUE)
alpha <- 0.05
pvalue
if (pvalue <= alpha) {
  cat("Reject the null hypothesis")
} else {
  cat("Fail to reject null hypothesis")
}

```
There is sufficient evidence to reject the null hypothesis and conclude that the slope coefficient on Education is less than -0.1 at the 0.05 significance level.

In addition to our data on Education and Income, we also have, for each individual, an indicator of
whether or not they are considered a member of the “Millenial” generation (defined as being born between
1980 and 1996).

d. (2 pts) Show code that runs a multiple regression with Income as the response variable, and Education,
Millenial, and the interaction between Education and Millenial as the predictor variables. We will
refer to this as Analysis II henceforth.

```{r}
edu$Millenial <- as.factor(edu$Millenial)
Analysis_II <- lm(Income ~ Education * Millenial, data = edu)
summary(Analysis_II)
```
e. (4 pts) Using the output from the regression, provide two separate prediction equations: one for
Millenial individuals, and one for Non-Millenial individuals.

The general equation including interaction terms is: 
$$\text{Income} = \beta_0 + \beta_1 \times \text{Education} + \beta_2 \times \text{Millenial} + \beta_3 \times (\text{Education} \times \text{Millenial})$$
For non-millenials individuals (Millenial = 0 which means Education x Millenial = 0):
$$ \text{Income}_{\text{non-Millenial}} = \beta_0 + \beta_1 \times \text{Education} $$
Substitute in the coefficients: $$\text{Income}_{\text{non-Millenial}} = 21.12971 + 0.39905 \times \text{Education}$$

For millenial individuals (when Millenial = 1, Education x Millenial = Education):
$$ \text{Income}_{\text{Millenial}} = \beta_0 + \beta_1 \times \text{Education} + \beta_2 + \beta_3 \times \text{Education} $$
Simplifying: $$\text{Income}_{\text{Millenial}} = (\beta_0 + \beta_2) + (\beta_1 + \beta_3) \times \text{Education}$$
Substitute in the coefficients: $$\text{Income}{\text{Millenial}} = (21.12971 - 8.15296) + (0.39905 - 0.08595) \times \text{Education}$$
$$\text{Income}{\text{Millenial}} = 12.97675 + 0.3131 \times \text{Education}$$

f. (4 pts) Consider the following two groups of individuals: group one is the collection of non-Millenials
in the rural area with 8 years of education, and group two is the collection of non-Millenial individuals
in the rural area with 7 years of education. Based on your analysis, is there evidence that the difference
in expected income between group one and group two is greater than $100 dollars? Conduct your test
at α = 0.05.

Ho: Difference in exp. income <= 100
Ha: Difference in exp. income > 100

Using this equation: $$\text{Income}{\text{Millenial}} = 12.97675 + 0.3131 \times \text{Education}$$ 
Difference in expected income between non-Millenials in the rural area with 8 years of education and non-Millenial individuals in the rural area with 7 years of education:
$$[21.12971 + 0.39905 \times 8] - [21.12971 + 0.39905 \times 7] = 0.39905 (thousandsof dollars) = 399.05 (dollars)$$
```{r}
tstat_f <- (399.05 - 100) / 0.14565
pvalue_f <- pt(tstat_f, df = Analysis_II$df.residual, lower.tail = FALSE)
alpha <- 0.05
pvalue_f
if(pvalue_f <= alpha) {
  cat("Reject the null hypothesis")
} else {
  cat("Fail to reject the null hypothesis")
}
```
Here we reject the null hypothesis and there's significant evidence that the difference in income is greater than $100.

g. (3 pts) Construct a 90% confidence interval for the difference between slope coefficients on Education in the population regression lines describing Millenials versus non-Millenials.

```{r}
coefficients_II <- coef(summary(Analysis_II))
interaction_coef <- coefficients_II["Education:MillenialTRUE", "Estimate"]
interaction_se <- coefficients_II["Education:MillenialTRUE", "Std. Error"]
df <- summary(Analysis_II)$df[2]
alpha_g <- 0.1
t_critical <- qt(1 - alpha_g / 2, df)
ci_lower <- interaction_coef - t_critical * interaction_se
ci_upper <- interaction_coef + t_critical * interaction_se
ci_lower
ci_upper
```
The 90% confidence interval for the difference between the slope coefficients on Education for Millenials versus non-Millenials is approximately: [(-0.41082, 0.23892)]

h. (2 pts) Is there evidence that the model yielding Analysis II provides an improvement in predictive
performance over a model that doesn’t use any predictor variables? Conduct your test at α = 0.05.

```{r}
null_model <- lm(Income ~ 1, data = edu)
anova <- anova(null_model, Analysis_II)
anova
pvalue_h <- anova$'Pr(>F)'[2]
alpha_h <- 0.05
pvalue_h
if(pvalue_h <= alpha) {
  cat("Reject the null hypothesis")
} else {
  cat("Fail to reject the null hypothesis")
}
```
Since we reject the null hypothesis, the model with predictor variables provides a significantly better fit than the intercept-only model at the 0.05 significance level.

i. (3 pts) A crotchety old man confronts you about these two analyses, and says that one of them must be incorrect. He says, “Analysis II tells me that if I went back to school for two more years I would earn more money, but Analysis I tells me that if I went back to school for two more years I would lose money. What a crock of [manure] - you must have [made a mistake somewhere]!” (his exact expression was censored for the purpose of this homework). What’s wrong with the crotchety old man’s interpretation of your analyses?

For Analysis I: The model indicates a negative relationship between education and income. This suggests that, on average, income decreases as education increases by 0.5172 for each additional year of education. Analysis I considers the entire sample without differentiating between subgroups.

For Analysis II: This model takes into account the interaction between Education and Millenial status. Analysis II distinguishes between Millenials and non-Millenials. It shows that within these subgroups, education indeed increases income, but the degree varies between Millenials and non-Millenials.

- For non-Millenials: Income increases by $399.05 per additional year of education.

- For Millenials: Income increases by $313.1 per additional year of education (0.39905 - 0.08595). 

So the old man is likely misinterpreting the results because he is applying the simple model's (Analysis I) finding across all contexts, without considering the richer, subgroup-specific findings (Analysis II). The interaction term in Analysis II indicates that the relationship between education and income differs for Millenials compared to non-Millenials, which is a critical insight missing in Analysis I.

j. (5 pts) To further convince the man of the error in his ways, you recall the advice from your Statistics professor that you must always plot your data. There is code provided in the R script accompanying this that creates a scatter plot of the data, but denotes Millenials and non-Millenials using different plotting symbols. The plot is also “jittered” to avoid overplotting (this means a small amount of noise is added to the x and y values being plotted). Study the plot. After reflection, explain in words the source of the differences between what you observed in Analysis I and Analysis II.
```{r}
Education <- edu$Education
Income <- edu$Income
Millenial= edu$Millenial
plot(Education, Income, type = "n", main="Scatter Plot", ylim = c(4, 40))
points(jitter(Education[Millenial==F],1), Income[Millenial == F], pch = 1, cex = 1)
points(jitter(Education[Millenial == T],1), Income[Millenial == T], pch = 17, cex = 1)
legend("topright", c("Non-Millenial", "Millenial"), pch = c(1, 17), cex = 1, bty = "n")
```
Observation:

Clusters of Points:

- Non-Millenials: These data points are scattered around different education levels with a certain trend in income

- Millenials: These data points look similarly scattered but might show a different trend compared to the non-Millenials

Trends:

- For Non-Millenials: Generally, as education increases, there's no significant upward or downward movement in income

- For Millenials: There might be a weaker upward trend in income with increasing education

Source of differences:

Analysis I fails to account for subgroup differences, aggregating all data points into a single trend. This overlooks key interaction effects. However, analysis II reveals more accurate subgroup-specific effects by considering the interaction term. By looking at the plot, we can see that non-Millenials and Millenials have different distributions and trends in their data. The scatter plot shows distinct clusters and trends for Millenials and non-Millenials, so we need the interaction term in Analysis II to capture these distinct patterns accurately.
